---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
unzip("activity.zip", exdir=".", overwrite=TRUE)
activity <- read.csv("activity.csv")
head(activity)
```


## What is mean total number of steps taken per day?

```{r}
totalDailySteps <- aggregate(steps ~ date, data = activity, sum)
hist(totalDailySteps$steps, breaks = 10, 
     xlab="Total Number of Steps Per Day",
     main = "Average Daily Number of Steps")
meanSteps <- mean(totalDailySteps$steps)
medianStep <- median(totalDailySteps$steps)
```
The average of the total number of steps taken per day is `r format(meanSteps, nsmall = 1)` the median is `r medianStep`.


## What is the average daily activity pattern?

```{r}
library(ggplot2)

intervalSteps <- aggregate(steps ~ interval, data = activity, mean)
ggplot(intervalSteps, aes(x=interval, y=steps)) +
    geom_line() +
    ggtitle("Average number of steps taken per 5-minute interval")

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
subset(intervalSteps, intervalSteps$steps == max(intervalSteps$steps, na.rm = TRUE))
```

## Imputing missing values

How many values are missing?

```{r}
sum(is.na(activity$steps))
```

How are the missing values distributed?

```{r}
missingSteps <- subset(activity, is.na(steps))
missingDates <- unique(missingSteps$date)
missingIntervalCount <- length(unique(missingSteps$interval))
```
There are `r length(missingDates)` dates and `r missingIntervalCount` intervals missing. Lets take a look at the dates:
```{r}
library(kableExtra)
kable(data.frame(date=missingDates))
```

Since there are full days missing, in one case even consecutive days I decided to replace all NA values with 0.

```{r}
activity$steps[is.na(activity$steps)] <- 0
```

Here is the updated histogram.
```{r}
totalDailySteps <- aggregate(steps ~ date, data = activity, sum)
hist(totalDailySteps$steps, breaks = 10, 
     xlab="Total Number of Steps Per Day",
     main = "Average Daily Number of Steps")
meanSteps <- mean(totalDailySteps$steps)
medianStep <- median(totalDailySteps$steps)
```
The new average of the total number of steps taken per day is `r format(meanSteps, nsmall = 1)` the median is `r format(medianStep, nsmall = 1)`. They are lower than before because of the added zeros.



## Are there differences in activity patterns between weekdays and weekends?
```{r message=FALSE}
library(chron)
library(ggplot2)
library(tidyr)

isWeekend <- is.weekend(as.Date(activity$date))

weekendSteps <- aggregate(steps ~ interval, data = activity[isWeekend,], mean)
weekdaySteps <- aggregate(steps ~ interval, data = activity[!isWeekend,], mean)
merged <- merge(weekdaySteps, weekendSteps, by = "interval")
names(merged) <- c("interval", "weekday", "weekend")
df <- pivot_longer(merged, contains("week"), names_to = "day",
                 values_to = "steps")
df$day <- as.factor(df$day)

ggplot(df, aes(x=interval, y=steps)) +
    geom_line() +
    facet_wrap(~day, ncol = 1) +
    ggtitle("Average number of steps taken per 5-minute interval across weekdays and weekends")
```


It seems that the activity on weekends is more evenly distributed during the day. The bulk of activity during weekdays happends in the morning.
